Response to reviewers V2

### 2025-10-15 paper_plos/paper_plos_V2.qmd
V2 revision, first draft

abstract: |
  Transportation planning has long prioritized the efficiency of movement. However, the concept of accessibility represents a more comprehensive evolution, shifting focus from movement to the potential to reach--or spatially interact with--desired destinations. Despite growing recognition of accessibility-based planning approaches, the concept remains fragmented, with inconsistent definitions and unclear interpretations. To this end, this paper offers a methodological contribution by specifying a family of accessibility measures grounded in the shared theoretical roots of gravity-based accessibility and spatial interaction models, particularly their balancing factors. From this foundation, we outline four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model). These measures can be interpreted as either the number of accessible opportunities or accessible population (i.e., market potential). A toy example illustrates how they produce interpretable, unit-based values, offering a clearer and more coherent basis for accessibility analysis.

@@

Accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability. Following this introduction, this paper revisits that lineage--from Ravenstein's Newtonian gravitational expression of spatial interaction [@ravensteinLawsMigration1889], through Hansen [@hansen1959] to Wilson [@wilson1971]--to propose a unified family of constrained accessibility measures. We illustrate each case of the family with a numerical toy example and conclude by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.

@@

# Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)

@@

Wilson's work is notable for many reasons. Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation. This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows. In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled. While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units.

As a preview for the work ahead: Wilson's balancing factors can be manipulated into _proportional allocation factors_ and equations can be established to reflect accessibility. The resulting $ij$ flows can be understood as accessibility flows--and aggregated at the zone level to express accessibility in units of actual opportunities or population.

In this way, these balancing factors can be used to recover what Stewart's $G$ was meant to accomplish, but through a different conceptual path. But before demonstrating a family of accessibility measures that are sensitive to constraints, in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge. This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this paper.

@@

In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based logic to the concept of accessibility. So, in the next section, we demonstrate how applying Wilson's framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns. This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity. This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially proposed in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).

@@

Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter. This section introduces this reframing: defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality. This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.

>>>>

abstract: |
  Transportation planning has long prioritized the efficiency of movement. However, the concept of accessibility represents a more comprehensive evolution, shifting focus from movement (i.e., trips) to the potential to spatially interact with desired destinations. Despite growing recognition of accessibility-based planning approaches, the concept remains fragmented, with inconsistent definitions and unclear interpretations. To this end, this paper offers a methodological contribution by specifying a family of accessibility measures that are grounded in the shared 'gravity-based' theoretical roots of spatial interaction models, particularly their balancing factors. From this foundation, we outline four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model). These measures can be interpreted as either the number of accessible opportunities or accessible population (i.e., market potential). A toy example illustrates how they produce interpretable, unit-based values, offering a clearer and more coherent basis for accessibility analysis.

@@

The remainder of this paper proceeds as follows. Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and @careyPrinciplesSocialScience1858 (Section 2), moving through early researchers like @ravensteinLawsMigration1885 to @stewartDemographicGravitationEvidence1948 who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in @hansen1959 became the dominant approach in planning practice (Section 4). Section 5 presents the entropy-based family of spatial interaction models in @wilson1971, explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates. Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations. Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied--unconstrained, singly-constrained, and doubly-constrained cases--demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices. We illustrate each measure with numerical examples that clarify the practical implications of different constraint assumptions. Section 8 concludes by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.

The key insight is that accessibility need not be unit-inconsistent. By building from spatial interaction theory's constraint framework rather than the gravitational analogy alone, we show how to construct accessibility measures that are simultaneously grounded in behavioural principles (the gravitational analogy) as well as being expressed in interpretable units (opportunities or population) that are sensitive to the known to the region through empirically-set constraints.

@@

# Early research on human spatial interaction: from Ravenstein (1885) to Stewart (1948)

@@

Wilson's work is notable for many reasons. Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation. This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows. The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @Wilson1971) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally. In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled. While Wilson's model produces results that are in units of flow tethered to the system of analysis, which facilitates such interpretations, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units.

Before demonstrating the derivation of family of accessibility measures that are sensitive to constraints (Section 7), in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge. This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this paper.

@@

In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based logic to the concept of accessibility. So, in the next section, we demonstrate how applying Wilson's framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns. This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity. This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially suggested in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).

@@

Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter. This section introduces this reframing: defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility ($\grave{a}$ la Hansen [@hansen1959]) from a relationship describing proportionality to equality. This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.

### 2025-10-15 paper_plos/paper_plos_V2.tex
V2 revision, first draft

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement.
However, the concept of accessibility represents a more comprehensive
evolution, shifting focus from movement to the potential to reach--or
spatially interact with--desired destinations. Despite growing
recognition of accessibility-based planning approaches, the concept
remains fragmented, with inconsistent definitions and unclear
interpretations. To this end, this paper offers a methodological
contribution by specifying a family of accessibility measures grounded
in the shared theoretical roots of gravity-based accessibility and
spatial interaction models, particularly their balancing factors. From
this foundation, we outline four memebers of the family: the
`unconstrained' measure (i.e., Hansen-type accessibility), the `total
constrained' measure (i.e., a constrained version of the Hansen-type
accessibility), the `singly constrained' measure (i.e., related to the
popular two step floating catchment approach - 2SFCA), and the `doubly
constrained' measure representing realized access (i.e., equal to the
doubly constrained spatial interaction model). These measures can be
interpreted as either the number of accessible opportunities or
accessible population (i.e., market potential). A toy example
illustrates how they produce interpretable, unit-based values, offering
a clearer and more coherent basis for accessibility analysis.

\linenumbers

\section{Introduction}\label{introduction}

@@

Accessibility and spatial interaction modelling literature share common
headwaters, and the latter has given careful attention to measurement
units and their interpretability. Following this introduction, this
paper revisits that lineage--from Ravenstein's Newtonian gravitational
expression of spatial interaction \citep{ravensteinLawsMigration1889},
through Hansen \citep{hansen1959} to Wilson \citep{wilson1971}--to
propose a unified family of constrained accessibility measures. We
illustrate each case of the family with a numerical toy example and
conclude by discussing how these constrained measures can inform
planning decisions and improve clarity in accessibility analysis.

@@

\section{Early research on human spatial interaction: from Ravenstein
(1889) to Stewart
(1948)}\label{early-research-on-human-spatial-interaction-from-ravenstein-1889-to-stewart-1948}

@@

Wilson's work is notable for many reasons. Rather than relying on a
universal constant like \(G\) or a scaling factor to balance units,
Wilson's models calibrate interaction flows through known empirical
constraints using principles of entropy maximisation. This results in
interpretable, balanced (given the constraints), and unit-consistent
\(ij\) flows. In this way, spatial interaction modeling tradition can be
seen to have succeeded where accessibility modeling stalled. While
Wilson's model produces results that are in units of flow tethered to
the system of analysis, Hansen-type measures (still widely used today in
accessibility work) yield outputs that reflect proportional -- not equal
-- relationships and typically lack interpretable units.

As a preview for the work ahead: Wilson's balancing factors can be
manipulated into \emph{proportional allocation factors} and equations
can be established to reflect accessibility. The resulting \(ij\) flows
can be understood as accessibility flows--and aggregated at the zone
level to express accessibility in units of actual opportunities or
population.

In this way, these balancing factors can be used to recover what
Stewart's \(G\) was meant to accomplish, but through a different
conceptual path. But before demonstrating a family of accessibility
measures that are sensitive to constraints, in the next section we
review how conceptually intertwined the spatial interaction modelling
and accessibility literatures are, and where they began to diverge. This
investigation sheds light on why accessibility research may have failed
to adopt a comparable approach focused, until this paper.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_V1_files/figure-pdf/fig-docs-per-year-1.pdf}

@@

In sum, despite the interpretative advantaged offered by the statistical
logic of Wilson \citep{wilson1971}'s framework, neither the spatial
interaction literature that cites Hansen \citep{hansen1959} nor the
accessibility literature citing Wilson \citep{wilson1971} has
meaningfully applied Wilson's constraint-based logic to the concept of
accessibility. So, in the next section, we demonstrate how applying
Wilson's framework enables accessibility to move from a proportional
relationship to calibrated equality--tying outputs to tangible system
knowns. This re-expression of accessibility using constraints yields
interpretable, unit-consistent measures of opportunity. This approach
takes the same path of entropy-maximisation as in Wilson
\citep{wilson1971}, and does not rely on specifying some universal
constant \(G\) like initially proposed in Stewart
\citep{stewartDemographicGravitationEvidence1948} (recall,
Eq~\ref{eq-stewart-population-potential-sum} which Hansen
\citep{hansen1959} operationalised).

@@

Despite the close conceptual ties between accessibility and spatial
interaction modeling, the former has not meaningfully absorbed the
constraint-based logic of the latter. This section introduces this
reframing: defining a family of accessibility measures using Wilson's
approach grounded in statistical mechanics, shifting place-based
accessibility (a la Hansen \citep{hansen1959}) from a relationship
describing proportionality to equality. This shift addresses the
fundamental issue of interpretability associated with the proportional
nature of Hansen-type accessibility indicators we previous outlined.

>>>>

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement.
However, the concept of accessibility represents a more comprehensive
evolution, shifting focus from movement (i.e., trips) to the potential
to spatially interact with desired destinations. Despite growing
recognition of accessibility-based planning approaches, the concept
remains fragmented, with inconsistent definitions and unclear
interpretations. To this end, this paper offers a methodological
contribution by specifying a family of accessibility measures that are
grounded in the shared `gravity-based' theoretical roots of spatial
interaction models, particularly their balancing factors. From this
foundation, we outline four memebers of the family: the `unconstrained'
measure (i.e., Hansen-type accessibility), the `total constrained'
measure (i.e., a constrained version of the Hansen-type accessibility),
the `singly constrained' measure (i.e., related to the popular two step
floating catchment approach - 2SFCA), and the `doubly constrained'
measure representing realized access (i.e., equal to the doubly
constrained spatial interaction model). These measures can be
interpreted as either the number of accessible opportunities or
accessible population (i.e., market potential). A toy example
illustrates how they produce interpretable, unit-based values, offering
a clearer and more coherent basis for accessibility analysis.

\linenumbers
\section{Introduction}\label{introduction}

@@

The remainder of this paper proceeds as follows. Sections 2-4 trace the
historical development of spatial interaction and accessibility
research, beginning with Newtonian gravitational analogies and
\citet{careyPrinciplesSocialScience1858} (Section 2), moving through
early researchers like \citet{ravensteinLawsMigration1885} to
\citet{stewartDemographicGravitationEvidence1948} who theorised and
formalized spatial interaction patterns (Section 3), and examining how
the ``gravity-based'' accessibility approach in \citet{hansen1959}
became the dominant approach in planning practice (Section 4). Section 5
presents the entropy-based family of spatial interaction models in
\citet{wilson1971}, explaining how the introduction of ``constraints''
(based on top-down known information as part of entropy-maximisation)
produce interpretable, unit-consistent flow estimates. Section 6
explores why accessibility and spatial interaction literature diverged
into separate branches despite their shared conceptual foundations.
Section 7 constitutes the paper's core contribution: we derive a family
of accessibility measures corresponding to different constraints
applied--unconstrained, singly-constrained, and doubly-constrained
cases--demonstrating how each yields zone-level accessibility values
expressed in meaningful units (opportunities or population) rather than
unit-inconsistent indices. We illustrate each measure with numerical
examples that clarify the practical implications of different constraint
assumptions. Section 8 concludes by discussing how these constrained
measures can inform planning decisions and improve clarity in
accessibility analysis.

The key insight is that accessibility need not be unit-inconsistent. By
building from spatial interaction theory's constraint framework rather
than the gravitational analogy alone, we show how to construct
accessibility measures that are simultaneously grounded in behavioural
principles (the gravitational analogy) as well as being expressed in
interpretable units (opportunities or population) that are sensitive to
the known to the region through empirically-set constraints.

@@

\section{Early research on human spatial interaction: from Ravenstein
(1885) to Stewart
(1948)}\label{early-research-on-human-spatial-interaction-from-ravenstein-1885-to-stewart-1948}

@@

Wilson's work is notable for many reasons. Rather than relying on a
universal constant like \(G\) or a scaling factor to balance units,
Wilson's models calibrate interaction flows through known empirical
constraints using principles of entropy maximisation. This results in
interpretable, balanced (given the constraints), and unit-consistent
\(ij\) flows. The balancing factors themselves have been subject to
various interpretations--as terminal costs
\citep{dieter1962distribution}, weighted mean values
\citep{kirbyNormalizingFactorsGravity1970}, or as accessibility measures
themselves \citep{cesario1977new} (as suggested in \citet{Wilson1971})
and as rents \citep{morphetGravityModelCalibration2023}--reflecting
ongoing efforts to understand what these mathematical constructs
represent behaviourally. In this way, the spatial interaction modeling
tradition can be seen to have succeeded where accessibility modeling
stalled. While Wilson's model produces results that are in units of flow
tethered to the system of analysis, which facilitates such
interpretations, Hansen-type measures (still widely used today in
accessibility work) yield outputs that reflect proportional -- not equal
-- relationships and typically lack interpretable units.

Before demonstrating the derivation of family of accessibility measures
that are sensitive to constraints (Section 7), in the next section we
review how conceptually intertwined the spatial interaction modelling
and accessibility literatures are, and where they began to diverge. This
investigation sheds light on why accessibility research may have failed
to adopt a comparable approach focused, until this paper.

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_V2_files/figure-pdf/fig-docs-per-year-1.pdf}

@@

In sum, despite the interpretative advantaged offered by the statistical
logic of Wilson \citep{wilson1971}'s framework, neither the spatial
interaction literature that cites Hansen \citep{hansen1959} nor the
accessibility literature citing Wilson \citep{wilson1971} has
meaningfully applied Wilson's constraint-based logic to the concept of
accessibility. So, in the next section, we demonstrate how applying
Wilson's framework enables accessibility to move from a proportional
relationship to calibrated equality--tying outputs to tangible system
knowns. This re-expression of accessibility using constraints yields
interpretable, unit-consistent measures of opportunity. This approach
takes the same path of entropy-maximisation as in Wilson
\citep{wilson1971}, and does not rely on specifying some universal
constant \(G\) like initially suggested in Stewart
\citep{stewartDemographicGravitationEvidence1948} (recall,
Eq~\ref{eq-stewart-population-potential-sum} which Hansen
\citep{hansen1959} operationalised).

@@

Despite the close conceptual ties between accessibility and spatial
interaction modeling, the former has not meaningfully absorbed the
constraint-based logic of the latter. This section introduces this
reframing: defining a family of accessibility measures using Wilson's
approach grounded in statistical mechanics, shifting place-based
accessibility (\(\grave{a}\) la Hansen \citep{hansen1959}) from a
relationship describing proportionality to equality. This shift
addresses the fundamental issue of interpretability associated with the
proportional nature of Hansen-type accessibility indicators we previous
outlined.
